---
title: "Two-Way Comparison Figures with ggcompare()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-Way Comparison Figures with ggcompare()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Every computed chunk needs datarium (the example data) and emmeans (the
# pooled-model simple comparisons). Skip gracefully if either is absent so the
# vignette still builds.
has_deps <- requireNamespace("datarium", quietly = TRUE) &&
  requireNamespace("emmeans", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96,
  eval = has_deps
)
```

```{r not-available, echo = FALSE, eval = !has_deps, results = "asis"}
cat("> This vignette needs the **datarium** and **emmeans** packages, which are",
    "not installed, so its output is not shown here.")
```

A common publication figure compares an outcome across **two** categorical
factors at once -- for example a score across *gender* and *education level* --
and annotates it with the interaction from a two-way ANOVA plus pairwise
comparisons within each group. `ggcompare()` assembles this figure in a single
call.

## The one-call two-way figure

Map the first factor to `x` and the second factor to `color` (or `fill`). That
second mapping switches `ggcompare()` into its two-way mode.

```{r two-way}
library(ggpubr)
library(rstatix)

data("jobsatisfaction", package = "datarium")

ggcompare(
  jobsatisfaction,
  x = "gender", y = "score", color = "education_level"
)
```

The figure carries:

- **dodged boxes** for the two factors, with jittered points and group means;
- **pairwise comparison brackets** of the education levels *within each gender*;
- a **subtitle** naming the two-way ANOVA interaction; and
- a **caption** describing the pairwise test.

In two-way mode the pairwise `method` defaults to `"emmeans_test"` (simple
comparisons from the pooled model), `p.adjust.method` to `"bonferroni"`, and
`hide.ns` to `TRUE`. Each of these is a default only -- pass the argument to
override it.

## It matches the by-hand analysis

The annotation reproduces the numbers you would compute by hand. The subtitle is
the interaction row of the two-way ANOVA:

```{r anova}
jobsatisfaction %>% anova_test(score ~ gender * education_level)
```

and the brackets are the pooled-model emmeans comparisons within each gender:

```{r pwc}
jobsatisfaction %>%
  group_by(gender) %>%
  emmeans_test(score ~ education_level, p.adjust.method = "bonferroni") %>%
  select(gender, group1, group2, p.adj, p.adj.signif)
```

## Customizing

Every element is overridable. For instance, compare the *genders within each
education level* instead (swap the comparison direction with `pwc.group.by`),
switch the base geometry, and show non-significant brackets:

```{r customize}
ggcompare(
  jobsatisfaction,
  x = "gender", y = "score", color = "education_level",
  base = "violin",
  pwc.group.by = "legend.var",
  hide.ns = FALSE
)
```

You can also compare each group to a reference level (`ref.group`), choose a
different pairwise `method`/`p.adjust.method`, or pass extra styling through
`pwc.args`, `add.params` and `...` (forwarded to the base builder). See
`?ggcompare` for the full argument list.

## Naming an effect on any plot

The subtitle machinery is also available directly through `add_test_label()`.
Given a plot with `x` and a second factor, `group.by` computes the two-way
ANOVA and `effect` chooses which effect to report -- the interaction (default),
a named term, or `"all"` for a multi-line label of both main effects and the
interaction:

```{r add-test-label}
p <- ggboxplot(jobsatisfaction, x = "gender", y = "score",
  color = "education_level")

add_test_label(p, group.by = "education_level", effect = "all", type = "text")
```
